---
title: "Reproducible Research: Peer Assessment 1"
author: "R Zach Lamberty"
date: "08/16/2014"
output: html_document
---

## Loading and preprocessing the data

Load this sucker (assuming that this is being knit from within the same
directory as "activity.zip" is stored):
```{r}
activity <- read.csv(unz("activity.zip", "activity.csv"))
```

Looks good 'nuff for me.

## What is mean total number of steps taken per day?

Because some entire days have no information regarding the number of steps, we
will filter out all NA before ddply-ing rather than in the `mean` and `sum`
functions (thus avoiding pesky `NaN`s)
```{r}
library(plyr)
notNA.activity <- activity[!is.na(activity$steps), ]
activity.by.day <- ddply(
  .data=notNA.activity,
  .variables=.(date),
  .fun=function(x) {
    c(
      total.steps=sum(x$steps),
      mean.steps=mean(x$steps),
      median.steps=median(x$steps)
    )
  }
)
```

Hisssssssssssstogram
```{r}
library(ggplot2)
ggplot(activity.by.day, aes(x=total.steps)) +
  geom_histogram(color="black", fill="blue") +
  xlab("Steps") +
  ggtitle("Total steps taken per day")
```

All right, all right, I'll report everything -- no need to be so `mean`.
```{r, echo=FALSE}
activity.by.day[, -2]
```

Though it looks odd at first, I believe the all-0 medians to be accurate -- the
vast majority of the time people are simply not taking steps.

## What is the average daily activity pattern?

Now we are more interested in the activity day-to-day, so let's re-ddply
ourselves.
```{r}
activity.by.interval <- ddply(
  .data=activity,
  .variables=.(interval),
  .fun=function(x) {
    c(
      ave.steps=mean(x$steps, na.rm=T),
      max.steps=max(x$steps, na.rm=T)
    )
  }
)
```

Hot to plot 
```{r}
ggplot(activity.by.interval, aes(x=interval, y=ave.steps)) +
  geom_line(color="blue", fill="blue") +
  ylab("Steps") +
  ggtitle("Average number of steps per interval")
```

Taking it to the max,
```{r}
max.all <- max(activity.by.interval$max.steps)
max.ave <- max(activity.by.interval$ave.steps)
interval.max.all <- activity.by.interval[activity.by.interval$max.steps == max.all, "interval"]
interval.max.ave <- activity.by.interval[activity.by.interval$ave.steps == max.ave, "interval"]
interval.diff <- interval.max.ave - interval.max.all
interval.as.time <- function(i) {
  x <- as.character(i)
  paste(
    substr(x, 1, nchar(x) - 2),
    substr(x, nchar(x) - 1, nchar(x)),
    sep=":"  
  )
}
```

To answer the specific question asked in this project description, the interval
which has the largest average number of steps (`r max.ave`) during it across all
days is interval `r interval.as.time(interval.max.ave)`. For what it is worth,
this is over `r interval.as.time(interval.diff)` away from the maximum number of
steps (`r max.all`) taken in a **single** interval, taken on a single day within
interval `r interval.as.time(interval.max.all)`. Looks like some early bird is
gettin that worm.

## Imputing missing values

In the activity data set, there are

```{r}
nrow(activity) - sum(complete.cases(activity))
```

rows with missing data in them.

We will replace all missing step data with the average number of steps in the
corresponding interval over all days which have data:

```{r}
activity.imputed <- activity
for (i in 1:nrow(activity.by.interval)) {
  interval <- activity.by.interval[i, "interval"]
  aveSteps <- activity.by.interval[i, "ave.steps"]
  toFix <- activity.imputed$interval == interval & is.na(activity.imputed$steps)
  activity.imputed[toFix, "steps"] <- aveSteps
}
```

After this step we have
```{r}
nrow(activity.imputed) - sum(complete.cases(activity.imputed))
```

missing data points in our `activity.imputed` data frame.

Now, repeating the steps in part 1 with this `activity.imputed` data frame;

```{r}
activity.imputed.by.day <- ddply(
  .data=activity.imputed,
  .variables=.(date),
  .fun=function(x) {
    c(
      total.steps=sum(x$steps),
      mean.steps=mean(x$steps),
      median.steps=median(x$steps)
    )
  }
)
```

Hisssssssssssstogram part deux
```{r}
ggplot(activity.imputed.by.day, aes(x=total.steps)) +
  geom_histogram(color="black", fill="blue") +
  xlab("Steps") +
  ggtitle("Total steps taken per day")
```

All right, all right, I'll report everything *again* -- no need to be so `mean`.
```{r, echo=FALSE}
activity.imputed.by.day[, -2]
```

You'll notice that we have introducted some weird behavior here in that several
days now have `34.11` median steps. These dates are the same for which we just
imputed all `NA` step values from the average over all days in the corresponding
interval. What we are reporting on those rows is not really the mean number of
steps on *those* days, but the mean number of steps on *any* day.

We would have to do considerably more work (*i.e.* generate an appropriate
distribution) to create values for steps on the "`NA` days" that had the *any*
day mean but a median of 0.

Finally, a side-by-side comparison thanks to our friends at `ggplot2`:
```{r}
df.join <- rbind(
  cbind(activity.by.day, dataset="with na"),
  cbind(activity.imputed.by.day, dataset="imputed")
)
df.join$dataset <- as.factor(df.join$dataset)
ggplot(df.join, aes(x=total.steps, fill=dataset)) +
  geom_histogram(color="black", position="dodge") +
  xlab("Steps") +
  ggtitle("Total steps taken per day")
```

From this side-by-side it's pretty easy to see that what we did was replace the
8 `NA` days with the median behavior -- in this case, 8 more counts of the
median number of steps per day in the **imputed** data set than in the
**with na** data set.

## Are there differences in activity patterns between weekdays and weekends?

We start by creating our factor:
```{r}
activity.imputed$datetype <- factor(
  x=weekdays(as.Date(activity.imputed$date)) %in% c("Saturday", "Sunday"),
  levels=c(FALSE, TRUE),
  labels=c("weekday", "weekend")
)
```

Creating the splittable factor is the same as above in step 3:
```{r}
activity.imputed.by.interval <- ddply(
  .data=activity.imputed,
  .variables=.(interval, datetype),
  .fun=function(x) {
    c(
      ave.steps=mean(x$steps, na.rm=T),
      max.steps=max(x$steps, na.rm=T)
    )
  }
)
```

Now, I think the side-by-side plot is much better for comparison than the panel
plot:
```{r}
ggplot(activity.imputed.by.interval,
       aes(x=interval, y=ave.steps, color=datetype)) +
  geom_line() +
  ylab("Steps") +
  ggtitle("Average number of steps per interval")
```

The story here is pretty clear: on weekdays, people are taking many more steps
early -- this is their commute to work. This activity all happens up until about
9:30 AM, at which point the trend totally inverts, and people are far more
sedantary on week- / work-days than they are on the weekends. There is a brief
reversal again at quittin' time (c. 6PM) in which work commuters are walking a
bit more than their weekend counterparts, but this tails off again once they
get home.

And for those of you who simply cannot live without a panel plot, I give you
```{r}
ggplot(activity.imputed.by.interval, aes(x=interval, y=ave.steps)) +
  geom_line(color="blue") +
  facet_wrap(~ datetype) +
  ylab("Steps") +
  ggtitle("Average number of steps per interval")
```